Drop Electro-Bouncing in Low-Gravity

### 2018-07-04 doc/manuscript/manuscript.tex
minor formatting changes

\documentclass[12pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{fontspec} % This line only for XeLaTeX and LuaLaTeX
\usepackage{pgfplots}
\usepackage{pgf}
\usepackage{bookmark}
\usepackage{color}
\usepackage{tikz}
\graphicspath{ {../figures/} }
%\usepackage{svg}
\usepackage{verbatim}
\usepackage{color,soul}
\usepackage{listings}
\usepackage{setspace}
\usepackage[caption=false]{subfig}
\newlength\figureheight
\newlength\figurewidth
\setlength\figureheight{7cm}
\setlength\figurewidth{10cm}
\usepackage[
backend=biber,
style=phys,
sorting=none
]{biblatex}
\addbibresource{thesis.bib}
\usepackage{import}
\let\pgfimageWithoutPath\pgfimage 
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[#1]{../figures/#2}}
\usepackage{array} 
\usepackage{appendix}
\usepackage{makeidx}
\makeindex

\title{\textsf{\textbf{Drop Electro-Bouncing in Low-Gravity}}}
\vspace{-25mm}
\author{Erin S. Schmidt, Mark M. Weislogel}
\date{}

@@

\centering %\vspace*{-10ex}
{\large \textsf{\textbf{Electro-Drop Bouncing in Low-Gravity}}}

@@

\tableofcontents
\afterpage{\blankpage}
\listoffigures
\afterpage{\blankpage}

@@

Our terrestrially born intuition about how liquids flow is easily confounded in a low-gravity environment. This should come as little surprise, as we are creatures evolved at the bottom of a gravity well; gravity is natural to us. However when the magnitude of gravitational body forces become small, other forces come into play in fluid dynamical problems, which are otherwise negligible under normal circumstances in 1-$g_0$. Our 1-$g_0$ cognitive bias then leads to a plethora of problems which are both simultaneously trivially easy to solve in 1-$g_0$, and which despite decades of study continue to elude solutions in a low-gravity context. One example of these problems is the so-called ``phase separation'' problem of separating a gas phase from a multi-phase flow (or the reverse case), without the aid of buoyancy due to gravity. Sans buoyancy, otherwise mundane activities such as venting a gas or settling a liquid in a tank become problematic \cite{petrash_controlling_1964}. Trapped bubbles can vapor lock life-support, power, or propulsion systems \cite{jenson_passive_2014}. Issues of phase separation have so bedeviled human endeavors in space that an entire Apollo Saturn 1B mission (AS-203) was earmarked to study them \cite{hastings_saturn_1965}. This problem has for some time motivated a quest for substitute body forces, and the present work follows in that well-worn tradition.

A solid beginning to rectifying our problem of poor low-gravity intuition, would be to quantify how wrong our intuition is. First, let us clear up a common misnomer about the state of ``zero-gravity''. The theory of general relativity is based on a fundamental equivalence of gravity and acceleration. By measuring acceleration alone, one cannot determine whether one is in gravity-free space or free-falling in a gravitational field (or conversely whether or not one is uniformly accelerating, or standing at rest in a gravitational field). This is sometimes called the equivalence principle of general relativity (e.g. the equivalence of gravitational and inertial mass). In theoretical terms, a `zero-g' experiment performed in a drop tower on the surface of the Earth is equivalent to the same experiment performed in the nearly gravity-free deep space between galaxies, despite the fact that the drop tower itself is enveloped in a 1-$g_0$ gravity field, where $g_0=9.81$ m/s$^2$ is the standard gravitational acceleration at sea-level. At the orbital altitude of the International Space Station (ISS) the gravity level is still nearly 90\% of its value at sea-level on Earth, but the astronauts aboard ISS feel weightlessness because they are in a state of free-fall. In a practical sense true zero-g does not exit anywhere in the universe as there are always small, but nonetheless, measurable forces which can accelerate any massive body. For instance, in low-Earth orbit accelerations due to aerodynamic resistance in the tenuous outer atmosphere are on the order of $1 \times 10^{-6}$ $g_0$ (from hence comes the term ``microgravity''). This small drag acceleration provides some ersatz gravity, but there are also other forces present. Foremost of these are capillary forces.

Capillary forces are due to the cumulative effects of a large number of very short range molecular interactions and in a 1-$g_0$ context are usually quite weak, relatively speaking. However, in a low-gravity environment other forces are so weak that the capillary forces becomes extraordinarily strong in a relative sense. In comparing the relative magnitudes of gravity and the capillary force we arrive at our figure of merit for assessing how wrong our fluid mechanical intuition is likely to be in a low-gravity context. This figure of merit is the Bond number dimensionless group
\[ \mathbb{B}\mbox{o} \equiv \frac{g R^2 \Delta \rho}{\gamma}, 
\]
where $g$ is the acceleration, $R$ is the characteristic interfacial length scale, $\Delta \rho$ is the difference in densities across the interface (which simplifies to $\rho$, for large $\Delta \rho$, as in the case of a air-water interface), and where $\gamma$ is the surface tension. In cases where $\mathbb{B}\mbox{o}$ is small a liquid is effectively in low-gravity regardless of the nominal local gravity level. Very small water drops ($R < 1$ mm), for instance, are in low-gravity for all intents and purposes and have nearly spherical shapes. Contrarily, for cases of large $R$ in a space environment, such as with large diameter spacecraft propellant tanks then $ \mathbb{B}\mbox{o} \gg 1$ even though the acceleration remains small. This is actually a quite useful result, for on the basis of dimensional similarity it is possible to accurately simulate low-gravity fluid mechanics (say of large propellant tanks) even in small drop towers simply by scaling dimensionless groups like the Bond number.

@@

\section{Spontaneous Drop Jump}
Low-gravity phase separation, drop dynamics, and the dynamics and applications of the so-called spontaneous drop or puddle jump, is an active area of study at the Portland State University Dryden Drop Tower laboratory (DDT). When a nonwetting, gravity-dominated sessile drop (e.g. a puddle), which is initially at rest on a surface in the Cassie-Baxter state suddenly undergoes a large step reduction in $\mathbb{B}\mbox{o}$ it will spontaneously jump away from the surface. The Cassie-Baxter state, sometimes also known as the Fakir state, is a metastable heterogenously non-wetting state on a textured surface characterized by the existence of a vapor-layer that separates roughness features of the surface from the bottom of the drop. The spontaneous drop jump was first observed experimentally in the Soviet Union by Kirko \emph{et al.} \cite{kirko_phenomenon_1970} in 1970 for drops of mercury in hydrochloric acid, and later by Wollman \emph{et al.} in 2016 for water drops in air in a set of experiments conducted using drop towers \cite{wollman_more_2016}. The kinetic energy of the jump is supplied by the defect in free surface energy as the new minimum energy surface equilibrium has approximately constant curvature. If the energy losses by viscous and contact line dynamics are neglected, the available kinetic energy is approximated by
\[KE_2 = SE_2 - SE_1 = [(\gamma A)_{gl} + (\gamma A)_{sg}]_2 - [(\gamma A)_{ls} + (\gamma A)_{gl} + (\gamma A)_{sg}]_1, \]
where $KE$ and $SE$ are the respective kinetic and surface energies, $A$ is the surface area, the subscripts $1$ and $2$ denote the intial 1-$g_0$, and low-g states, and the subscripts ${gl}$, ${sg}$, and ${ls}$ denote the gas-liquid, solid-gas and liquid-solid interfaces respectively. The motive force is due to the inertia of internal flows which occur as the interface deflects under the suddenly lessened $\mathbb{B}\mbox{o}$. The characteristic time scale of the rolling up scales as $t_j \sim (\rho V_d/\sigma)^{1/2}$ \cite{attari_puddle_2016}, which resembles the contact time, $\tau \approx 2.6(\rho R^3_d/\sigma)^{1/2}$ presented by Richard \emph{et al.} in 2002 for the related problem of drops impacting hydrophobic surfaces in 1-$g_0$ \cite{richard_surface_2002}. For drops with radial symmetry and sufficiently high initial $\mathbb{B}\mbox{o}$, inward radial capillary waves coalesce at the axis, leading to geysering and creation of satellite droplets by the Rayleigh-Plateau breakup of the geyser. In the case of smaller jumping drops the capillary waves are gradually damped by viscous forces. A time-lapsed composite photograph of spontaneous drop jump can be seen in Figure \ref{fig:nobounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{drop_jump}
\caption{The trajectory of a 0.2 mL spontaneous drop jump is captured in a composite image over a period $\sim 0.67$ s presented at $\approx 10$ Hz. \label{fig:nobounce}}
\end{figure}

@@

We begin with some preliminary observations of the phenomenon:
\begin{itemize}
\item Observed maximum drop (de-)accelerations are on the order of $\sim$30 cm/s$^2$ for a range of drop volumes $0.03 \lessapprox V_d \lessapprox 0.5$ mL.
\item The water drops are attracted to regions of high electric field. The horizontal (surface plane parallel) components of the drop trajectory usually oscillate about their initial position during the experiment (except in cases of nearly pure 1-D vertical translation). For especially small drops close to the spontaneous drop jump limit ($V_d \sim$ 0.01 mL) the drops do not jump, but translate across the surface in a rolling regime until either they reach a local maximum of the electric field, or until their motion is sufficiently damped by contact line hysteresis where pinning ultimately arrests their motion.
\item The drops appear to have net free charge. In cases of multiple simultaneous drop jumps the drops repel each other as they bounce or roll in orbital motion around regions of high field.
\item The magnitude of the drop trajectory maxima (apoapses) appear to be related to the drop volume (mass), and initial jump velocity (inertia), and to the electric field strength.
\end{itemize}

There are several possible origins of the electric charge that are ultimately responsible for this phenomenon. It is well known that water acquires positive free charge when in contact with certain polymers, especially polytetrafluoroethylene (PTFE) \cite{langmuir_surface_1938}, through a process called contact charging. PTFE, on the other hand, tends to readily acquire negative charge by contact with water. The superhydrophobic surfaces used in the spontaneous drop jump experiments have thin (nanometric) PTFE coatings, and we observe that it is extremely easy to produce significant surface potentials $\varphi_s \sim$ 100-500 V by simply flowing streams of distilled water over them. A study of this water on PTFE contact charging phenomenon was conducted by Yatsuzuka \emph{et al.} \cite{yatsuzuka_electrification_1994}, who suggest that this process results from formation of an electrical double layer driven by selective adsorption of ($\mbox{OH}^-$) ions at the polymer surface; other recent work strongly supports this hypothesis \cite{beattie_intrinsic_2006, strazdaite_water_2015}. Yatsuzuka \emph{et al.} also found that the specific charge on the drops in contact with the surface depends on their both surface velocity and conductivity. The contact charging mechanism is the most likely accidental source of charge on the suphydrophobic surfaces used in our work. Given the large roughness, or the ratio of projected to actual surface area, of the superhydrophobic surfaces used in the experiment, and given that the drops are initially in a Cassie-Baxter state, a somewhat electrically resistive air layer is maintained that reduces grounding of the drops despite the large potential difference between them and the surface charges. If this electrostatic potential is greater than the release of surface energy under the sudden change in $\mathbb{B}\mbox{o}$, in the form of the drop kinetic energy, we expect that the drop will return to an equilibrium state on the charged surface which minimizes the potential.

@@

\section{Applications}
This study of the electro-drop bounce phenomenon may have results which can be extended to useful applications in a space environment. External surfaces of spacecraft tend to become charged with time due to the space environment itself. Surface charging largely occurs due to low energy electrons (3-50 keV) which do not penetrate the surface of the spacecraft external structure \cite{czepiela_charging_1997}. The ultimate sources of these charges are trapped charged particles of the van Allen radiation belts, galactic cosmic rays, and the solar wind. This deposited charge accumulates and can lead to significant potential differences between different parts of a spacecraft, sometimes leading to breakdown and discharge called Paschen discharge. Deep dielectric charging occurs when higher energy charged particles penetrate the surface of a dielectric material, which can also lead to large potential differences if the dielectric leakage is lower than the external charging rate. In a fluid mechanical context, this accumulated charge could be problematic during any multiphase venting process, as might be expected to occur during autonomous satellite servicing refueling operations, or during boil-off venting of cryogens stored over long periods in a space environment (as in many crewed Mars mission scenarios, or in cryogenic propellant depot operations). Such venting, if occurring in a region of strong field, will also tend to produce a stream of charged drops, which will be decelerated by the field, causing their eventual return, impacting and potentially contaminating the spacecraft surfaces.

As previously noted, robust phase separation is critical to high reliability multiphase systems used in thermal control and life support. Active solid-state phase separation for life support multiphase flows, especially high void-fraction droplet disperse flows is a possible application of this work. Such flows are encountered in dish washing, laundry, waste solids drying, food processing, the Sabatier CO$_2$ reaction, and possibly in vapor-compression cycle condensers. Phase separation for other disperse droplet flows include electrostatic droplet separators for high-efficiency Rankine-cycle turbines, which require a droplet-free vapor phase entering the turbine, but where conventional superheat approaches come with severe mass penalties \cite{unterberg_zero_1962}. More speculatively, results of this work could be applied in high temperature liquid droplet radiators with electrostatic collection \cite{white_liquid_1987}. Removal of satellite droplets produced during pipetting in wet-lab research outside of a glovebox environment aboard ISS has also been recently suggested as a application of the work [private communication, NASA JSC, E. Unger, 2017]. Drops can become spontaneously charged by contact with standard micropipette tips \cite{choi_spontaneous_2013}, and this free charge can possibly be leveraged for the purposes of phase separation in low-gravity.

@@

Modeling the electrostatic force is more involved, but we begin with the standard electrohydrodynamic (EHD) approximation \cite{saville_electrohydrodynamics:_1997}. Under a DC electric field, we assume that the real part of the dielectric permittivity $\epsilon$, $\mbox{Re} \langle \epsilon \rangle \approx  \mbox{constant}$. We also assume that electric currents are small enough that the effects of magnetic fields can be neglected. The validity of this assumption rests on the characteristic time scale $\tau_e = \epsilon /\sigma_e \ll 1$, where $\tau_e$ is the ratio of absolute dielectric permittivity $\epsilon = \kappa \epsilon_0$, to conductivity $\sigma_e$, of the medium, $\kappa$ is the relative dielectric permittivity, and $\epsilon_0$ is the vacuum permittivity. This characteristic time $\tau_e$ is also known as the relaxation time, and is a measure of how quickly the polarization of a dielectric responds to a change in electric field. Given the conductivity and permittivity in the limiting case of extremely-pure water ($ \epsilon \sim 80$, $\sigma_e = 18.2 \times 10^{6}$ $\Omega\mbox{cm}$) \cite{yatsuzuka_electrification_1994}, we estimate $\tau_e \approx 4 \times 10^{-6}$ s. The relaxation time for the common distilled water that is actually used in the drop experiments is undoubtedly shorter due to the presence of solvated ions. Neglecting the effects of an electric double layer on hydration of ions in the water or the ambient atmosphere due to the relatively massive size of the drops studied, the assumption of small relaxation time further implies that the free charge present in the drops will remain approximately constant during the typical time interval of a low-gravity experiment.

@@

\section{Scaling}
For small DEP forces the equation of motion Equation \ref{gov_eqn_subs} is non-linear non-homogeneous and must be solved numerically. Nevertheless we are curious if we can obtain approximate solutions to the equation analytically using perturbation methods. This is aided by non-dimensionalizing the governing equations, forming dimensionless groups, and determining whether any are particularly small or large, and whether any fortuitous opportunities for simplification appear. Introducing the scaled variables
\begin{equation}
 t^* = \frac{t}{t_c}, \hspace{10 mm} y^* = \frac{y}{y_c}, \end{equation}
where $y_c$ and $t_c$ are characteristic length and time scales respectively, and using the coordinate transformation $y(0) - R = 0$, the governing equation becomes
\begin{eqnarray}
& {y^*}'' = - \frac{1}{2} \frac{C_D \rho A y_c}{m} {y^*}'^2
- \frac{q E_0 t_c^2}{m y_c} E^* ( {y^*} ) 
- \frac{k q^2 t_c^2}{16 \pi \epsilon_0 R^2 m y_c} \left( \frac{y_c}{R} {y^*} + 1 \right)^{-2}, & \label{scaled_eqn}
\end{eqnarray}
subject to 
\begin{equation*}
{y^*}(0) = 0, \hspace{5 mm} \mbox{and} \hspace{5 mm} {y^*}'(0) = \frac{U_0 t_c}{y_c} .
\end{equation*}
From Equation \ref{scaled_eqn} we note the existence of several dimensionless groups
\[ \mathbf{\Pi}_1 = \frac{C_D \rho A y_c}{2 m}, \hspace{5 mm}
\mathbf{\Pi}_2 = \frac{q E_0 t_c^2}{m y_c}, \hspace{5 mm}
\mathbf{\Pi}_3 = \frac{k q^2 t_c^2}{16 \pi \epsilon_0 R^2 m y_c}, \hspace{5 mm}
\mathbf{\Pi}_4 = \frac{y_c}{R}, \hspace{5 mm}
\mathbf{\Pi}_5 = \frac{U_0 t_c}{y_c},\]
such that Equation \ref{scaled_eqn} becomes
\begin{equation}
\label{pi_terms}
 {y^*}'' = - \mathbf{\Pi}_1 {y^*}'^2
- \mathbf{\Pi}_2 E^* ( {y^*} ) 
- \mathbf{\Pi}_3 \left( \mathbf{\Pi}_4 {y^*} + 1 \right)^{-2},
\end{equation}
subject to
\begin{equation*}
{y^*}(0) = 0, \hspace{5 mm} \mbox{and} \hspace{5 mm} {y^*}'(0) = \mathbf{\Pi}_5 .
\end{equation*}

>>>>

\documentclass[12pt,a4paper,oneside]{book}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
%\usepackage{fontspec} % This line only for XeLaTeX and LuaLaTeX
\usepackage{pgfplots}
\usepackage{pgf}
\usepackage{bookmark}
\usepackage{color}
\usepackage{tikz}
\graphicspath{ {../figures/} }
%\usepackage{svg}
\usepackage{verbatim}
\usepackage{color,soul}
\usepackage{listings}
\usepackage{setspace}
\usepackage[caption=false]{subfig}
\newlength\figureheight
\newlength\figurewidth
\setlength\figureheight{7cm}
\setlength\figurewidth{10cm}
\usepackage[
backend=biber,
style=phys,
sorting=none
]{biblatex}
\addbibresource{thesis.bib}
\usepackage{import}
\let\pgfimageWithoutPath\pgfimage 
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[#1]{../figures/#2}}
\usepackage{array} 
\usepackage{appendix}
\usepackage{makeidx}
\usepackage{sectsty}
%\chapternumberfont{\large} 
%\chaptertitlefont{\large}
%\allsectionsfont{\normalsize}
\makeindex

\title{Drop Electro-Bouncing in Low-Gravity}
\vspace{-25mm}
\author{Erin S. Schmidt, Mark M. Weislogel}
\date{}

@@

\centering %\vspace*{-10ex}
{\textsf{{Electro-Drop Bouncing in Low-Gravity}}}

@@

\tableofcontents
%\afterpage{\blankpage}
\listoffigures
%\afterpage{\blankpage}

@@

Our terrestrially born intuition about how liquids flow is easily confounded in a low-gravity environment. This should come as little surprise, as we are creatures evolved at the bottom of a gravity well; gravity is natural to us. However when the magnitude of gravitational body forces become small, other forces come into play in fluid dynamical problems, which are otherwise negligible under normal circumstances in 1-$g_0$, where $g_0=9.81$ m/s$^2$ is the standard gravitational acceleration at sea-level. Our 1-$g_0$ cognitive bias then leads to a plethora of problems which are both simultaneously trivially easy to solve in 1-$g_0$, and which despite decades of study continue to elude solutions in a low-gravity context. One example of these problems is the so-called ``phase separation'' problem of separating a gas phase from a multi-phase flow (or the reverse case), without the aid of buoyancy due to gravity. Sans buoyancy, otherwise mundane activities such as venting a gas or settling a liquid in a tank become problematic \cite{petrash_controlling_1964}. Trapped bubbles can vapor lock life-support, power, or propulsion systems \cite{jenson_passive_2014}. Issues of phase separation have so bedeviled human endeavors in space that an entire Apollo Saturn 1B mission (AS-203) was earmarked to study them \cite{hastings_saturn_1965}. This problem has for some time motivated a quest for substitute body forces, and the present work follows in that well-worn tradition.

A solid beginning to rectifying our problem of poor low-gravity intuition, would be to quantify how wrong our intuition is. First, let us clear up a common misnomer about the state of ``zero-gravity''. The theory of general relativity is based on a fundamental equivalence of gravity and acceleration. By measuring acceleration alone, one cannot determine whether one is in gravity-free space or free-falling in a gravitational field (or conversely whether or not one is uniformly accelerating, or standing at rest in a gravitational field). This is sometimes called the equivalence principle of general relativity (e.g. the equivalence of gravitational and inertial mass). In theoretical terms, a `zero-g' experiment performed in a drop tower on the surface of the Earth is equivalent to the same experiment performed in the nearly gravity-free deep space between galaxies, despite the fact that the drop tower itself is enveloped in a 1-$g_0$ gravity field. At the orbital altitude of the International Space Station (ISS) the gravity level is still nearly 90\% of its value at sea-level on Earth, but the astronauts aboard ISS feel weightlessness because they are in a state of free-fall. In a practical sense true zero-g does not exist anywhere in the universe as there are always small, but nonetheless, measurable forces which can accelerate any massive body. For instance, in low-Earth orbit accelerations due to aerodynamic resistance in the tenuous outer atmosphere are on the order of $1 \times 10^{-6}$ $g_0$ (from hence comes the term ``microgravity''). This small drag acceleration provides some ersatz gravity, but there are also other forces present. Foremost of these are capillary forces.

Capillary forces are due to the cumulative effects of a large number of very short range molecular interactions and in a 1-$g_0$ context are usually quite weak, relatively speaking. However, in a low-gravity environment other forces are so weak that capillary forces become extraordinarily strong in a relative sense. In comparing the relative magnitudes of gravity and the capillary force we arrive at our figure of merit for assessing how wrong our fluid mechanical intuition is likely to be in a low-gravity context. This figure of merit is the Bond number dimensionless group
\[ \mathbb{B}\mbox{o} \equiv \frac{g R^2 \Delta \rho}{\gamma}, 
\]
where $g$ is the acceleration, $R$ is the characteristic interfacial length scale, $\Delta \rho$ is the difference in densities across the interface (which simplifies to $\rho$, for large $\Delta \rho$, as in the case of an air-water interface), and where $\gamma$ is the surface tension. In cases where $\mathbb{B}\mbox{o}$ is small a liquid is effectively in low-gravity regardless of the nominal local gravity level. Very small water drops ($R < 1$ mm), for instance, are in low-gravity for all intents and purposes and have nearly spherical shapes. Contrarily, for cases of large $R$ in a space environment, such as with large diameter spacecraft propellant tanks then $ \mathbb{B}\mbox{o} \gg 1$ even though the acceleration remains small. This is actually a quite useful result, for on the basis of dimensional similarity it is possible to accurately simulate low-gravity fluid mechanics (say of large propellant tanks) even in small drop towers simply by scaling dimensionless groups like the Bond number.

@@

\section{Spontaneous Drop Jump}
Low-gravity phase separation, drop dynamics, and the dynamics and applications of the so-called spontaneous drop or puddle jump, is an active area of study at the Portland State University Dryden Drop Tower laboratory (DDT). When a nonwetting, gravity-dominated sessile drop (e.g. a puddle), which is initially at rest on a surface in the Cassie-Baxter state suddenly undergoes a large step reduction in $\mathbb{B}\mbox{o}$ it will spontaneously jump away from the surface. The Cassie-Baxter state, sometimes also known as the Fakir state, is a metastable heterogenously non-wetting state on a textured surface characterized by the existence of a vapor-layer that separates roughness features of the surface from the bottom of the drop. The spontaneous drop jump was first observed experimentally in the Soviet Union by Kirko \emph{et al.} \cite{kirko_phenomenon_1970} in 1970 for drops of mercury in hydrochloric acid, and later by Wollman \emph{et al.} in 2016 for water drops in air in a set of experiments conducted using drop towers \cite{wollman_more_2016}. The kinetic energy of the jump is supplied by the defect in free surface energy as the new minimum energy surface equilibrium has approximately constant curvature. If the energy losses by viscous and contact line dynamics are neglected, the available kinetic energy is approximated by
\[KE_2 = SE_2 - SE_1 = [(\gamma A)_{gl} + (\gamma A)_{sg}]_2 - [(\gamma A)_{ls} + (\gamma A)_{gl} + (\gamma A)_{sg}]_1, \]
where $KE$ and $SE$ are the respective kinetic and surface energies, $A$ is the surface area, the subscripts $1$ and $2$ denote the intial 1-$g_0$, and low-g states, and the subscripts ${gl}$, ${sg}$, and ${ls}$ denote the gas-liquid, solid-gas and liquid-solid interfaces respectively. The motive force is due to the inertia of internal flows which occur as the interface deflects under the suddenly lessened $\mathbb{B}\mbox{o}$. The characteristic time scale of the rolling up of the contact line scales as $t_j \sim (\rho V_d/\sigma)^{1/2}$ \cite{attari_puddle_2016}, which resembles the contact time, $\tau \approx 2.6(\rho R^3_d/\sigma)^{1/2}$ presented by Richard \emph{et al.} in 2002 for the related problem of drops impacting hydrophobic surfaces in 1-$g_0$ \cite{richard_surface_2002}. For drops with radial symmetry and sufficiently high initial $\mathbb{B}\mbox{o}$, inward radial capillary waves coalesce at the axis, leading to geysering and creation of satellite droplets by the Rayleigh-Plateau breakup of the geyser. In the case of smaller jumping drops the capillary waves are gradually damped by viscous forces. A time-lapsed composite photograph of spontaneous drop jump can be seen in Figure \ref{fig:nobounce}.
\begin{figure}[htb]
\centering
\includegraphics[width=0.75\textwidth]{drop_jump}
\caption{The trajectory of a 0.2 mL spontaneous drop jump is captured in a composite image over a period $\sim 0.67$ s presented at $\approx 10$ Hz. \label{fig:nobounce}}
\end{figure}

@@

We begin with some preliminary observations of the phenomenon:
\begin{itemize}
\item Observed maximum drop (de-)accelerations are on the order of $\sim$30 cm/s$^2$ for a range of drop volumes $0.03 \lessapprox V_d \lessapprox 0.5$ mL.
\item The water drops are attracted to regions of high electric field. The horizontal (surface plane parallel) components of the drop trajectory usually oscillate about some central position during the experiment (except in cases of nearly pure 1-D vertical translation). For especially small drops close to the spontaneous drop jump limit ($V_d \sim$ 0.01 mL) the drops do not jump, but translate across the surface in a rolling regime until either they reach a local maximum of the electric field, or until their motion is sufficiently damped by contact line hysteresis where pinning ultimately arrests their motion.
\item The drops appear to have net free charge. In cases of multiple simultaneous drop jumps the drops repel each other as they bounce or roll in orbital motion around regions of high field.
\item The magnitude of the drop trajectory maxima (apoapses) appear to be related to the drop volume (mass), and initial jump velocity (inertia), and to the electric field strength.
\end{itemize}

There are several possible origins of the electric charge that are ultimately responsible for this phenomenon. It is well known that water acquires positive free charge when in contact with certain polymers, especially polytetrafluoroethylene (PTFE), through a process called contact charging \cite{langmuir_surface_1938}. PTFE, on the other hand, tends to readily acquire negative charge by contact with water. The superhydrophobic surfaces used in the spontaneous drop jump experiments have thin (nanometric) PTFE coatings, and we observe that it is extremely easy to produce significant surface potentials $\varphi_s \sim$ 100-500 V by simply flowing streams of distilled water over them. A study of this water on PTFE contact charging phenomenon was conducted by Yatsuzuka \emph{et al.} \cite{yatsuzuka_electrification_1994}, who suggest that this process results from formation of an electrical double layer driven by selective adsorption of ($\mbox{OH}^-$) ions at the polymer surface; other recent work strongly supports this hypothesis \cite{beattie_intrinsic_2006, strazdaite_water_2015}. Yatsuzuka \emph{et al.} also found that the specific charge on the drops in contact with the surface depends on their both surface velocity and conductivity. The contact charging mechanism is the most likely accidental source of charge on the suphydrophobic surfaces used in our work. Given the large roughness, or the ratio of projected to actual surface area, of the superhydrophobic surfaces used in the experiment, and given that the drops are initially in a Cassie-Baxter state, a somewhat electrically resistive air layer is maintained that reduces grounding of the drops despite the large potential difference between them and the surface charges. If this electrostatic potential is greater than the release of surface energy under the sudden change in $\mathbb{B}\mbox{o}$, in the form of the drop kinetic energy, we expect that the drop will return to an equilibrium state on the charged surface which minimizes the potential.

@@

\section{Applications}
This study of the electro-drop bounce phenomenon may have results which can be extended to useful applications in a space environment. External surfaces of spacecraft tend to become charged with time due to the space environment itself. Surface charging largely occurs due to low energy electrons (3-50 keV) which do not penetrate the external surface of the spacecraft \cite{czepiela_charging_1997}. The ultimate sources of these charges are trapped charged particles of the van Allen radiation belts, galactic cosmic rays, and the solar wind. This deposited charge accumulates and can lead to significant potential differences between parts of a spacecraft, sometimes leading to breakdown and discharge called Paschen discharge. Deep dielectric charging occurs when higher energy charged particles penetrate the surface of a dielectric material, which can also lead to large potential differences if the dielectric leakage is lower than the external charging rate. In a fluid mechanical context, this accumulated charge could be problematic during any multiphase venting process, as might be expected to occur during autonomous satellite servicing refueling operations, or during boil-off venting of cryogens stored over long periods in a space environment (as in many crewed Mars mission scenarios, or in cryogenic propellant depot operations). Such venting, if occurring in a region of strong field, will also tend to produce a stream of charged drops, which will be decelerated by the field, causing their eventual return, impacting and potentially contaminating the spacecraft surfaces.

As previously noted, robust phase separation is critical to high reliability multiphase systems used in thermal control and life support. Active solid-state phase separation for life support multiphase flows, especially high void-fraction droplet disperse flows is a possible application of this work. Such flows are encountered in dish washing, laundry, waste solids drying, food processing, the Sabatier CO$_2$ reaction, and possibly in vapor-compression cycle condensers. Phase separation for other droplet disperse flows include electrostatic droplet separators for high-efficiency Rankine-cycle turbines, which require a droplet-free vapor phase entering the turbine, but where conventional superheat approaches come with severe mass penalties \cite{unterberg_zero_1962}. More speculatively, results of this work could be applied in high temperature liquid droplet radiators with electrostatic collection \cite{white_liquid_1987}. Removal of satellite droplets produced during pipetting in wet-lab research outside of a glovebox environment aboard ISS has also been recently suggested as a application of the work [private communication, NASA JSC, E. Unger, 2017]. Drops can become spontaneously charged by contact with standard micropipette tips \cite{choi_spontaneous_2013}, and this free charge can possibly be leveraged for the purposes of phase separation in low-gravity.

@@

Modeling the electrostatic force is more involved, but we begin with the standard electrohydrodynamic (EHD) approximation \cite{saville_electrohydrodynamics:_1997}. Under a DC electric field, we assume that the real part of the dielectric permittivity $\epsilon$, $\mbox{Re} \left( \epsilon \right) \approx  \mbox{constant}$. We also assume that electric currents are small enough that the effects of magnetic fields can be neglected. The validity of this assumption rests on the characteristic time scale $\tau_e = \epsilon /\sigma_e \ll 1$, where $\tau_e$ is the ratio of absolute dielectric permittivity $\epsilon = \kappa \epsilon_0$, to conductivity $\sigma_e$, of the medium, $\kappa$ is the relative dielectric permittivity, and $\epsilon_0$ is the vacuum permittivity. This characteristic time $\tau_e$ is also known as the relaxation time, and is a measure of how quickly the polarization of a dielectric responds to a change in electric field. Given the conductivity and permittivity in the limiting case of extremely-pure water ($ \epsilon \sim 80$, $\sigma_e = 18.2 \times 10^{6}$ $\Omega\mbox{cm}$) \cite{yatsuzuka_electrification_1994}, we estimate $\tau_e \approx 4 \times 10^{-6}$ s. The relaxation time for the common distilled water that is actually used in the drop experiments is undoubtedly shorter due to the presence of solvated ions. Neglecting the effects of an electric double layer on hydration of ions in the water or the ambient atmosphere due to the relatively massive size of the drops studied, the assumption of small relaxation time further implies that the free charge present in the drops will remain approximately constant during the typical time interval of a low-gravity experiment.

@@

\section{Scaling}
For small DEP forces the equation of motion Equation \ref{gov_eqn_subs} is non-linear, non-homogeneous and must be solved numerically. Nevertheless we are curious if we can obtain approximate solutions to the equation analytically using perturbation methods. This is aided by non-dimensionalizing the governing equations, forming dimensionless groups, and determining whether any are particularly small or large, and whether any fortuitous opportunities for simplification appear. Introducing the scaled variables
\begin{equation}
 t^* = \frac{t}{t_c}, \hspace{10 mm} y^* = \frac{y}{y_c}, \end{equation}
where $y_c$ and $t_c$ are characteristic length and time scales respectively, and using the coordinate transformation $y(0) - R = 0$, the governing equation becomes
\begin{eqnarray}
& {y^*}'' = - \frac{1}{2} \frac{C_D \rho A y_c}{m} {y^*}'^2
- \frac{q E_0 t_c^2}{m y_c} E^* ( {y^*} ) 
- \frac{k q^2 t_c^2}{16 \pi \epsilon_0 R^2 m y_c} \left( \frac{y_c}{R} {y^*} + 1 \right)^{-2}, & \label{scaled_eqn}
\end{eqnarray}
subject to 
\begin{equation*}
{y^*}(0) = 0, \hspace{5 mm} \mbox{and} \hspace{5 mm} {y^*}'(0) = \frac{U_0 t_c}{y_c} .
\end{equation*}
From Equation \ref{scaled_eqn} we note the existence of several dimensionless groups
\[ \mathbf{\Pi}_1 = \frac{C_D \rho A y_c}{2 m}, \hspace{5 mm}
\mathbf{\Pi}_2 = \frac{q E_0 t_c^2}{m y_c}, \hspace{5 mm}
\mathbf{\Pi}_3 = \frac{k q^2 t_c^2}{16 \pi \epsilon_0 R^2 m y_c}, \hspace{5 mm}
\mathbf{\Pi}_4 = \frac{y_c}{R}, \hspace{5 mm}
\mathbf{\Pi}_5 = \frac{U_0 t_c}{y_c},\]
such that Equation \ref{scaled_eqn} becomes
\begin{equation}
\label{pi_terms}
 {y^*}'' = - \mathbf{\Pi}_1 {y^*}'^2
- \mathbf{\Pi}_2 E^* ( {y^*} ) 
- \mathbf{\Pi}_3 \left( \mathbf{\Pi}_4 {y^*} + 1 \right)^{-2},
\end{equation}
subject to
\begin{equation*}
{y^*}(0) = 0, \hspace{5 mm} \mbox{and} \hspace{5 mm} {y^*}'(0) = \mathbf{\Pi}_5 .
\end{equation*}
